---
title: "Lab week 11 - h2o"
subtitle: "Data Science and Machine Learning 2 - CEU 2018"
author: "Jeno Pal"
date: '2018-03-13'
output:
  html_document:
    df_print: paged
  html_notebook:
    df_print: paged
---

H2O: state-of-the-art machine learning software that is even
suitable for big datasets. 
It offers very efficient and scalable implementations of 
popular ML algorithms that can 

* run on distributed systems
* utilize multiple cores
* work with GPUs

Models estimated with h2o can be deployed to production environments
through Java objects (see [here](http://docs.h2o.ai/h2o/latest-stable/h2o-docs/productionizing.html)).
Also, h2o scales well compared to other competitor implementations
(see Szilard Pafka's famous benchmarks [here](https://github.com/szilard/benchm-ml)).

In general, best resource to learn is the
[documentation](http://docs.h2o.ai/h2o/latest-stable/h2o-docs/welcome.html) and many
tutorials are available on YouTube.

```{r}
library(data.table)
library(ISLR)
library(ggplot2)

library(h2o)
h2o.init()
```

```{r}
# stop it via h2o.shutdown()
```

```{r}
data <- data.table(Wage)
skimr::skim(data)
```

```{r}
data[, c("region","wage") := NULL]

h2o_data <- as.h2o(data)
str(h2o_data)
```

```{r}
splitted_data <- h2o.splitFrame(h2o_data, 
                                ratios = c(0.6, 0.2), 
                                seed = 123)
data_train <- splitted_data[[1]]
data_valid <- splitted_data[[2]]
data_test <- splitted_data[[3]]
```

GLM: __penalized__ linear regression methods.
```{r}
y <- "logwage"
X <- setdiff(names(h2o_data), c(y, "year"))

glm_fit <- h2o.glm(X, y, 
               training_frame = data_train,
               alpha = 0,
               nfolds = 5, 
               seed = 123)

# xval stands for 'cross-validation'
print(h2o.performance(glm_fit, xval = TRUE))
```

```{r}
print(h2o.rmse(glm_fit, xval = TRUE))
```

You can perform a grid search with cross-validation to tune hyperparamters.
```{r}
hyper_params <- list(alpha = c(0, .25, .5, .75, .1))

# build grid search with previously selected hyperparameters
glm_grid <- h2o.grid(x = X, 
                     y = y, 
                     training_frame = data_train, 
                     algorithm = "glm", 
                     lambda_search = TRUE,
                     nfolds = 5,
                     seed = 123,
                     hyper_params = hyper_params)
h2o.getGrid(grid_id = glm_grid@grid_id, sort_by = "rmse", decreasing = FALSE)
```

Random Forests are there for you to use as well.
```{r}
# random forests
rf_params <- list(ntrees = c(500),
                  mtries = c(2, 3, 5))

rf_grid <- h2o.grid(x = X, 
                    y = y, 
                    training_frame = data_train, 
                    algorithm = "randomForest", 
                    nfolds = 5,
                    seed = 123,
                    hyper_params = rf_params)

h2o.getGrid(grid_id = rf_grid@grid_id, sort_by = "rmse", decreasing = FALSE)
```

Just like GBMs.

```{r}
# GBM hyperparamters
gbm_params <- list(learn_rate = c(0.01, 0.05),
                    max_depth = c(2, 3, 5),
                    sample_rate = c(0.5),
                    col_sample_rate = c(0.5, 1.0))

# Train and validate a cartesian grid of GBMs
gbm_grid <- h2o.grid(x = X, 
                     y = y, 
                     training_frame = data_train, 
                     algorithm = "gbm", 
                     nfolds = 5,
                     seed = 123,
                     ntrees = 300,
                     hyper_params = gbm_params)

h2o.getGrid(gbm_grid@grid_id, sort_by = "rmse", decreasing = FALSE)
```

```{r}
# get best models
glm_model <- h2o.getModel(h2o.getGrid(glm_grid@grid_id)@model_ids[[1]])
rf_model <- h2o.getModel(h2o.getGrid(rf_grid@grid_id)@model_ids[[1]])
gbm_model <- h2o.getModel(h2o.getGrid(gbm_grid@grid_id)@model_ids[[1]])

# predict on validation set
validation_performances <- list(
  "glm" = h2o.rmse(h2o.performance(glm_model, newdata = data_valid)),
  "rf" = h2o.rmse(h2o.performance(rf_model, newdata = data_valid)),
  "gbm" = h2o.rmse(h2o.performance(gbm_model, newdata = data_valid))
)

validation_performances
```

```{r}
# test set performance
h2o.rmse(h2o.performance(gbm_model, newdata = data_test))
```

```{r}
# turn back h2oFrames to plain data.frame
prediction_vs_truth = data.table(cbind(
    as.data.frame(h2o.predict(gbm_model, newdata = data_test)),
    as.data.frame(data_test)[["logwage"]]
))

setnames(prediction_vs_truth, c("prediction", "truth"))


head(prediction_vs_truth)

ggplot(data = prediction_vs_truth,
       aes(x = truth, y = prediction)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(3, 6) + ylim(3, 6) +
  geom_point() +
  theme_minimal()
```

And, of course, there is deep learning, too! We'll come to that later.
Let's just try it out of the box and see validation set performance.
```{r}
dl_model <- h2o.deeplearning(X, y, 
                             training_frame = data_train,
                             reproducible = TRUE, # needed in deeplearning for full reproducibility
                             seed = 123)
h2o.rmse(h2o.performance(dl_model, newdata = data_valid))
```

XGboost is also available (however, not on Windows).
